Kaluza--Klein Cosmology from five-dimensional Lovelock--Cartan Theory

### 2016-11-14 KKosmology.tex
Second correction after the referee report.

The $N$-dimensional Lorentz curvature and torsion are given by the Cartan structure equations
\begin{align}
  \label{curvadef}
  \hat{R}^{AB} &= \mbox{d}\hat{\omega}^{AB}+\hat{\omega}^A_{\ \ C}\wedge\hat{\omega}^{CB} = \frac{1}{2} \hat{R}^{AB}{}_{CD} \; \hat{e}^C \wedge \hat{e}^D,\\
  \label{tordef}
  \hat{T}^A &= \mbox{d}\hat{e}^A+\hat{\omega}^A_{\ \ B}\wedge\hat{e}^B = \frac{1}{2} \hat{T}^{A}{}_{BC} \; \hat{e}^B \wedge \hat{e}^C. 
\end{align}
Using the definition of curvature in Eq.~\eqref{curvadef},  with the KK ansatz for the spin connection~\eqref{Nconnection}, we find
\begin{align}
  \label{R1}
  \hat{R}^{ab}&=R^{ab}+\sqrt{\phi}\alpha^{ab}F-\beta^a\wedge\beta^b\notag\\
  & \quad +\left(\mbox{D}\alpha^{ab}+\frac{1}{2}\alpha^{ab}\mbox{d}\ln\phi-2\beta^{[a}\gamma^{b]}\right)\wedge\hat{e}^N,\\
  \label{R2}
  \hat{R}^{Na}&=-\left(\mbox{D}\beta^a+\sqrt{\phi}\gamma^a F\right)\notag\\
  & \quad +\left(\alpha^a_{\ b}\beta^b-\mbox{D}\gamma^a-\frac{1}{2}\gamma^a\mbox{d}\ln\phi\right)\wedge\hat{e}^N,
\end{align}
where $R=\text{d}\omega+\omega\wedge\omega$ is the curvature of the reduced spacetime $M_{N-1}$. Similarly from the definition of torsion in Eq.~\eqref{tordef}, we find its distinctive parts to be~\footnote{Henceforth we will refer to as distinctive parts regarding the $M_{N-1}\times S^1$ decomposition.}
\begin{align}\label{T1}
  \hat{T}^a &= T^a+\left(\beta^a-\alpha^a_{\ \ b}e^b\right)\wedge\hat{e}^N,\\
  \label{T2}
  \hat{T}^N &= \sqrt{\phi}F-\beta_b\wedge e^b+\left(\frac{1}{2}\mbox{d}\ln\phi+\gamma_be^b\right)\wedge\hat{e}^N,
\end{align}
where $T=\text{d}e+\omega\wedge e$ is the torsion $2$-form of $M_{N-1}$.

@@

Considering the Bianchi identities for the KK structure described above, we find relevant information about the new fields. Taking the exterior covariant derivative over the $N$-dimensional curvature and torsion, the Bianchi identities are
\begin{equation}
  \hat{\text{D}}\hat{R}^{AB} = 0 \quad \text{and} \quad \hat{\text{D}}\hat{T}^A = \hat{R}^A_{\ B}\wedge\hat{e}^B.
\end{equation}
A careful decomposition of the first one into its distinctive parts, gives the Bianchi identity for the curvature of the reduced spacetime, together with the second derivative rules
\begin{equation}
  \begin{aligned}
    \text{D}R^{ab} &=0, & \text{D}\mbox{D}\alpha^{ab} &=R^a_{\ l}\alpha^{lb}+R^b_{\ l}\alpha^{al},\\
    \text{D}\text{D}\beta^a &= R^a_{\ b}\beta^{b}, & \text{D}\text{D}\gamma^a &= R^a_{\ b}\gamma^{b}.
  \end{aligned}
\end{equation}
This is taken as a proof of the tensorial nature of these new fields under the four-dimensional Lorentz transformations.~\footnote{The tensorial nature of these fields can be also derived from the decomposition of the transformation rule of $\hat{\omega}^{AB}$ under the Lorentz group.} The second Bianchi identity gives its equivalent for $M_{N-1}$, this is
\begin{equation*}
  \mbox{D}T^a=R^a_{\ b}\wedge e^b.
\end{equation*}

@@

In this work, we will focus on the region in the parameter space where
\begin{equation}
  \label{delta}
  \Delta\equiv\alpha_1^2-4\alpha_0\alpha_2\neq 0.
\end{equation}
This condition place us outside the Chern--Simons point. Otherwise, the field equations~\eqref{delta e} and~\eqref{delta w} are invariant under a larger gauge group ($AdS_5$), while the action~\eqref{action5EGB} becomes the Chern--Simons form for that group~\cite{Zanelli:2016cs,*Troncoso:1999pk}. Exact solutions with torsion in Chern--Simons gravity are given in~\cite{Aros:2007nn,Banados:2003cz}. Its holographic properties in the Riemannian case where studied in Ref.~\cite{Banados:2005rz} and in \cite{Banados:2006fe} when torsion is considered.

@@

%None of the other cases, $u_- = 0$ and $u_\pm < 0$, present bouncing behavior.
The allowed parameter space for $u_+$ and $u_-$ to be negative is shown in Fig.~\ref{cond_u_neg}, while the restriction $u_- = 0$ sets $\alpha_2 = \tfrac{1}{12 \alpha_0}$.~\footnote{Notice that in our choice of parameters $u_+$ cannot vanish.}

@@

As we can see from Eq.~\eqref{homotorsion}, $f$ corresponds to the completely antisymmetric part of torsion. Its solution given in Eq.~\eqref{f2sol} shows that it can be imaginary for certain periods of time. Additionally, it is the only function sensitive to the value of $k$. Thus, if a reality condition is imposed over $f$ for certain values of the coupling constants, it would provide a distinction criteria among open, flat and closed universes. In the light of this and given the experimental evidence supporting the flatness of the universe (see Ref.~\cite{Spergel:2006hy}) is that we restrict the discussion to the case where $k=0$.~\footnote{The cases where $k = \pm1$ can be studied analogously.} A reality condition over $f$ in Eq.~\eqref{f2sol} is satisfied within the region 
\begin{align*}
 & & 3\alpha_1 \pm \sqrt{6\Delta} &\geq 0 & &\mbox{and}& \alpha_2 &>0, \\
 & \mbox{or,} & 3\alpha_1 \pm \sqrt{6\Delta} &\leq 0 & &\mbox{and}& \alpha_2&<0.
\end{align*}
This should be taken into account to restrict the parameter region studied in Sec.~\ref{sec:sol}.  
A similar behavior for this function was found in Ref.~\cite{Toloza:2013wi}. Notice that the antisymmetric part of torsion is unseen by classical particles following geodesics and it does not couple minimally to spin-0 or spin-1 bosons. However, it appears as an effective interaction term for the Dirac Lagrangian in spacetimes with torsion~\cite{Hehl:1976kj}
\begin{align*}
  \mathcal{L}_{\text{Int}}&=-\frac{i}{8}T_{\alpha\mu\nu}\bar{\Psi}\Gamma^{\alpha\mu\nu}\Psi =\frac{3}{2}f\bar{\Psi}\Gamma^0\Gamma^5\Psi,
\end{align*}
which couples to the fermionic axial current. For imaginary values of $f(t)$, the authors in Ref.~\cite{Toloza:2013wi} argued that the loss of unitarity and the violation of the current conservation can be interpreted as particle creation.

@@

\begin{acknowledgments}
  The authors would like to thank A.~R.~Zerwekh, I.~Schmidt,
  S.~Kovalenko, N.~A.~Neill, C.~O.~Dib, O.~Miskovic, R.~Olea,
  R.~Gannouji, A.~Toloza, A.~Cisterna and J.~Zanelli for valuable
  comments on this work. \mbox{O.~C.-F.} has been partially supported
  by the PAI project No.~79140040 (CONICYT--Chile) and F.~R. by the
  Fondecyt project No.~1160423. The Centro Cient\'ifico Tecnol\'ogico
  de Valpara\'iso (CCTVal) is funded by the Chilean Government through
  the Centers of Excellence Basal Financing Program FS0821 of CONICYT.
\end{acknowledgments}
%_-----------------------------------------------------------------------------------------

@@

In Riemann--Cartan geometry, both the vielbein $e^A$ and the spin connection $\omega^{AB}$ are independent features of the manifold.~\footnote{Here, we will drop the hats since we will not deal with the KK decomposition.} The spin connection can be decomposed in a Riemannian part, which is metric dependent, $\tilde{\omega}^{AB}$ satisfying $\tilde{\mbox{D}}e^A=0$, and a contorsion piece $\kappa^{AB}=-\kappa^{BA}$, such that 
\begin{equation}\label{spin separation}
  \omega^{AB}=\tilde{\omega}^{AB}+\kappa^{AB}.
\end{equation} 
Therefore, the torsion 2-form defined as the covariant derivative of the vielbein with respect to the total spin connection, is
\begin{equation}
  T^A=\kappa^A_{\ B}\wedge e^B.
\end{equation}
On the other hand, the curvature $2$-form also suffers corrections with respect to the Riemannian one, due to the presence of torsional degrees of freedom. This can be seen explictly by taking the definition of curvature in Eq.~\eqref{curvadef} and~\eqref{spin separation} to find
\begin{equation}\label{curvature decomp}
  R^{AB} = \tilde{R}^{AB} + \tilde{\mbox{D}}\kappa^{AB} + \kappa^A_{\ C}\wedge\kappa^{CB},
\end{equation}
where $\tilde{R}^{AB}$ is the Riemannian curvature constructed with $\tilde{\omega}^{AB}$.

>>>>

The $N$-dimensional Lorentz curvature and torsion are given by the Cartan structure equations
\begin{align}
  \label{curvadef}
  \hat{R}^{AB} &= \mbox{d}\hat{\omega}^{AB}+\hat{\omega}^A_{\ \ C}\wedge\hat{\omega}^{CB} = \frac{1}{2} \hat{R}^{AB}{}_{CD} \; \hat{e}^C \wedge \hat{e}^D,\\
  \label{tordef}
  \hat{T}^A &= \mbox{d}\hat{e}^A+\hat{\omega}^A_{\ \ B}\wedge\hat{e}^B = \frac{1}{2} \hat{T}^{A}{}_{BC} \; \hat{e}^B \wedge \hat{e}^C. 
\end{align}
Using the definition of curvature in Eq.~\eqref{curvadef},  with the KK ansatz for the spin connection~\eqref{Nconnection}, we find
\begin{align}
  \label{R1}
  \hat{R}^{ab}&=R^{ab}+\sqrt{\phi}\alpha^{ab}F-\beta^a\wedge\beta^b\notag\\
  & \quad +\left(\mbox{D}\alpha^{ab}+\frac{1}{2}\alpha^{ab}\mbox{d}\ln\phi-2\beta^{[a}\gamma^{b]}\right)\wedge\hat{e}^N,\\
  \label{R2}
  \hat{R}^{Na}&=-\left(\mbox{D}\beta^a+\sqrt{\phi}\gamma^a F\right)\notag\\
  & \quad +\left(\alpha^a_{\ b}\beta^b-\mbox{D}\gamma^a-\frac{1}{2}\gamma^a\mbox{d}\ln\phi\right)\wedge\hat{e}^N,
\end{align}
where $R=\text{d}\omega+\omega\wedge\omega$ is the curvature of the reduced spacetime $M_{N-1}$. Similarly from the definition of torsion in Eq.~\eqref{tordef}, we find its distinctive parts to be\footnote{Henceforth we will refer to as distinctive parts regarding the $M_{N-1}\times S^1$ decomposition.}
\begin{align}\label{T1}
  \hat{T}^a &= T^a+\left(\beta^a-\alpha^a_{\ \ b}e^b\right)\wedge\hat{e}^N,\\
  \label{T2}
  \hat{T}^N &= \sqrt{\phi}F-\beta_b\wedge e^b+\left(\frac{1}{2}\mbox{d}\ln\phi+\gamma_be^b\right)\wedge\hat{e}^N,
\end{align}
where $T=\text{d}e+\omega\wedge e$ is the torsion $2$-form of $M_{N-1}$.

@@

Considering the Bianchi identities for the KK structure described above, we find relevant information about the new fields. Taking the exterior covariant derivative over the $N$-dimensional curvature and torsion, the Bianchi identities are
\begin{equation}
  \hat{\text{D}}\hat{R}^{AB} = 0 \quad \text{and} \quad \hat{\text{D}}\hat{T}^A = \hat{R}^A_{\ B}\wedge\hat{e}^B.
\end{equation}
A careful decomposition of the first one into its distinctive parts, gives the Bianchi identity for the curvature of the reduced spacetime, together with the second derivative rules
\begin{equation}
  \begin{aligned}
    \text{D}R^{ab} &=0, & \text{D}\mbox{D}\alpha^{ab} &=R^a_{\ l}\alpha^{lb}+R^b_{\ l}\alpha^{al},\\
    \text{D}\text{D}\beta^a &= R^a_{\ b}\beta^{b}, & \text{D}\text{D}\gamma^a &= R^a_{\ b}\gamma^{b}.
  \end{aligned}
\end{equation}
This is taken as a proof of the tensorial nature of these new fields under the four-dimensional Lorentz transformations.\footnote{The tensorial nature of these fields can be also derived from the decomposition of the transformation rule of $\hat{\omega}^{AB}$ under the Lorentz group.} The second Bianchi identity gives its equivalent for $M_{N-1}$, this is
\begin{equation*}
  \mbox{D}T^a=R^a_{\ b}\wedge e^b.
\end{equation*}

@@

In this work, we will focus on the region in the parameter space where
\begin{equation}
  \label{delta}
  \Delta\equiv\alpha_1^2-4\alpha_0\alpha_2 > 0.
\end{equation}
This condition place us outside the Chern--Simons point ($\Delta =
0$). Otherwise, the field equations~\eqref{delta e} and~\eqref{delta
  w} are invariant under a larger gauge group ($AdS_5$), while the
action~\eqref{action5EGB} becomes the Chern--Simons form for that
group~\cite{Zanelli:2016cs,*Troncoso:1999pk}. Exact solutions with
torsion in Chern--Simons gravity are given
in~\cite{Aros:2007nn,Banados:2003cz}. Holographic properties in the
Riemannian case where studied in Ref.~\cite{Banados:2005rz} and in
\cite{Banados:2006fe} when torsion is considered. Additionally, the
case $\Delta < 0$ has been consider in Ref.~\cite{Canfora:2013xsa}, as
well as its cosmology in Ref.~\cite{Canfora:2014iga}.

@@

%None of the other cases, $u_- = 0$ and $u_\pm < 0$, present bouncing behavior.
The allowed parameter space for $u_+$ and $u_-$ to be negative is shown in Fig.~\ref{cond_u_neg}, while the restriction $u_- = 0$ sets $\alpha_2 = \tfrac{1}{12 \alpha_0}$.\footnote{Notice that in our choice of parameters $u_+$ cannot vanish.}

@@

As we can see from Eq.~\eqref{homotorsion}, $f$ corresponds to the completely antisymmetric part of torsion. Its solution given in Eq.~\eqref{f2sol} shows that it can be imaginary for certain periods of time. Additionally, it is the only function sensitive to the value of $k$. Thus, if a reality condition is imposed over $f$ for certain values of the coupling constants, it would provide a distinction criteria among open, flat and closed universes. In the light of this and given the experimental evidence supporting the flatness of the universe (see Ref.~\cite{Spergel:2006hy}) is that we restrict the discussion to the case where $k=0$.\footnote{The cases where $k = \pm1$ can be studied analogously.} A reality condition over $f$ in Eq.~\eqref{f2sol} is satisfied within the region 
\begin{align*}
 & & 3\alpha_1 \pm \sqrt{6\Delta} &\geq 0 & &\mbox{and}& \alpha_2 &>0, \\
 & \mbox{or,} & 3\alpha_1 \pm \sqrt{6\Delta} &\leq 0 & &\mbox{and}& \alpha_2&<0.
\end{align*}
This should be taken into account to restrict the parameter region studied in Sec.~\ref{sec:sol}.  
A similar behavior for this function was found in Ref.~\cite{Toloza:2013wi}. Notice that the antisymmetric part of torsion is unseen by classical particles following geodesics and it does not couple minimally to spin-0 or spin-1 bosons. However, it appears as an effective interaction term for the Dirac Lagrangian in spacetimes with torsion~\cite{Hehl:1976kj}
\begin{align*}
  \mathcal{L}_{\text{Int}}&=-\frac{i}{8}T_{\alpha\mu\nu}\bar{\Psi}\Gamma^{\alpha\mu\nu}\Psi =\frac{3}{2}f\bar{\Psi}\Gamma^0\Gamma^5\Psi,
\end{align*}
which couples to the fermionic axial current. For imaginary values of $f(t)$, the authors in Ref.~\cite{Toloza:2013wi} argued that the loss of unitarity and the violation of the current conservation can be interpreted as particle creation.

@@

\begin{acknowledgments}
  The authors would like to thank A.~R.~Zerwekh, I.~Schmidt,
  S.~Kovalenko, N.~A.~Neill, C.~O.~Dib, O.~Miskovic, R.~Gannouji,
  A.~Toloza, A.~Cisterna and J.~Zanelli for valuable comments on this
  work. \mbox{O.~C.-F.} has been partially supported by the PAI
  project No.~79140040 (CONICYT--Chile) and F.~R. by the Fondecyt
  project No.~1160423. The Centro Cient\'ifico Tecnol\'ogico de
  Valpara\'iso (CCTVal) is funded by the Chilean Government through
  the Centers of Excellence Basal Financing Program FS0821 of CONICYT.
\end{acknowledgments}
%_-----------------------------------------------------------------------------------------

@@

In Riemann--Cartan geometry, both the vielbein $e^A$ and the spin connection $\omega^{AB}$ are independent features of the manifold.\footnote{Here, we will drop the hats since we will not deal with the KK decomposition.} The spin connection can be decomposed in a Riemannian part, which is metric dependent, $\tilde{\omega}^{AB}$ satisfying $\tilde{\mbox{D}}e^A=0$, and a contorsion piece $\kappa^{AB}=-\kappa^{BA}$, such that 
\begin{equation}\label{spin separation}
  \omega^{AB}=\tilde{\omega}^{AB}+\kappa^{AB}.
\end{equation} 
Therefore, the torsion 2-form defined as the covariant derivative of the vielbein with respect to the total spin connection, is
\begin{equation}
  T^A=\kappa^A_{\ B}\wedge e^B.
\end{equation}
On the other hand, the curvature $2$-form also suffers corrections with respect to the Riemannian one, due to the presence of torsional degrees of freedom. This can be seen explictly by taking the definition of curvature in Eq.~\eqref{curvadef} and~\eqref{spin separation} to find
\begin{equation}\label{curvature decomp}
  R^{AB} = \tilde{R}^{AB} + \tilde{\mbox{D}}\kappa^{AB} + \kappa^A_{\ C}\wedge\kappa^{CB},
\end{equation}
where $\tilde{R}^{AB}$ is the Riemannian curvature constructed with $\tilde{\omega}^{AB}$.
